Parsing Postfix log files

### 2007-12-02 doc/logparser.tex
Tested LMA again, got the code working this time. Improved the review and criticism, not that I have output to do something with.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in
a form suitable for a systems administrator to search using grep(1), awk(1)
and the myriad of other standard Unix text processing utilities already
available, producing a \CSV{} file which can subsequently be imported into
a MySQL or Berkeley DB database.  The decision to support both \CSV{} and
Berkeley DB appears to have been a limiting factor: \CSV{} is a very simple
format where a record is stored in a single line, with fields separated by
a comma or other punctuation symbol.  Problems with \CSV{} files include
the difficulty in escaping separators, providing multiple values for a
field (e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.

@@

The data stored is limited to time and date, hostname and \IP{} address of
client, sender and recipient addresses, and the \SMTP{} code.  Handling of
multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
may be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper, but the options appear to be:

    \item Discard some of the data; this option is obviously undesirable,
        and presumably was not chosen by the \LMA{} authors.

    \item Combine all data into a \CSV{} line, which is then embedded into
        the main \CSV{} line for that mail, making searching even more
        difficult and requiring a second separating character.

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the database schema into the
database and provide commands for displaying it.  Berkeley DB does not
embed a schema, as there is neither requirement nor benefit; it provides
\textbf{(key, value)} pairs, so any additional structuring of the data is
imposed by the application, thus it behoves the application to document
this structure.

\parsername{} is bundled with extensive documentation, including an
in-depth explanation of the database schema.  It supports multiple values
for every field for which multiple values are reasonable (e.g.\ the sender
address should have one value only, whereas the recipient address can have
multiple values).  The default database it uses is SQLite3, which supports
most standard \SQL{} features (its lacks full foreign key support), has
libraries available for most common programming languages, and a simple
text based client allowing ad-hoc queries with minimal effort.

XXX DISCUSS THE OUTPUT, OR LACK THEREOF, IN GREATER DETAIL, AFTER RERUNNING
\LMA{}\@.

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in sections~\refwithpage{complications} (initial
complications) and~\refwithpage{additional complications} (additional
complications).  The program fails to parse Postfix 2.2 or 2.3 log files;
indeed all it produces is a number of warning messages from the Perl
interpreter, without any other output, useful or otherwise.

\LMA{} does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT); the author was unable to parse any logs with \LMA{}, so
didn't have an opportunity to review the reports.  \parsername{} does not
produce any reports, it delegates that task to a separate, dedicated
program.

@@

>>>>

XXX THINK OF A NAME\@.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients), adding new fields,
and lack of a bundled schema describing the fields.

@@

The data stored is limited to time and date, hostname and \IP{} address of
client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
codes or remote servers\footnote{A single mail may be sent to multiple
remote servers if it was addressed to recipients in different domains, or
Postfix needs to try multiple servers for one or more recipients.} is not
explained in the paper; experimental observation shows that multiple
records are added when there are multiple recipients (sadly the records are
not associated in any way), and presumably the same approach is taken when
there are multiple destination servers.

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the database schema into the
database and provide commands for displaying it.  Berkeley DB does not
embed a schema, as there is neither requirement nor benefit; it only
provides \textbf{(key, value)} pairs, so any additional structuring of the
data is imposed by the application, thus it behoves the application to
document this structure.

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in sections~\refwithpage{complications} (initial
complications) and~\refwithpage{additional complications} (additional
complications).  It does not differentiate between different types of
rejections, so it is not suitable for the purposes of this project; the
data about which restriction caused the rejection is discarded, whereas an
important goal of this project is to retain that data to aid optimisation
and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
files generated on Solaris hosts because the fields added automatically
differ from those added on Linux hosts; logs from Solaris hosts (and
possibly other operating systems) thus require preprocessing before parsing
by \LMA{}.

\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
DAILY-REJECT initially failed with an error message from the Perl
interpreter; after making some corrections to the code it worked, and
produced the \CSV{} lines for the specified day where the \SMTP{} code
signifies a rejection.  All of these reports are trivially simple to
produce from the \CSV{} file using the standard Unix tool \texttt{awk},
e.g.\ the most complicated, DAILY-REJECT, is merely:

\begin{verbatim}
awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/ 
    { print $0; print " "; }' lma_output.txt
\end{verbatim}

Notes about the command above: there is an unnecessary line containing only
a single space output after each matching line, for compatibility with the
output of DAILY-REJECT\@.  Some records have an \SMTP{} code of 0, though
this is not a valid \SMTP{} code, so presumably this is due to \LMA{} not
parsing a line properly; \LMA{} considers this code to be a rejection, so
the command above must too.  In a tiny proportion of records the \SMTP{}
code is \textit{deferred\/} (again, not a valid code, and most likely due
to mis-parsing), and is treated as a rejection by \LMA{}, so the equivalent
\texttt{awk} command must do so too.

There are a number of differences in the output from DAILY-REJECT and the
\texttt{awk} command:

    \item \LMA{} does not escape or replace the separators used in the
        \CSV{} records when they are present in sender or recipient
        addresses, leading to some mangled records; these records are not
        present in the output from \texttt{awk}.

    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
        produces only one output line, whereas the \texttt{awk} command
        produces all matching lines.

@@

\glossary{name={Solaris},description={
    Solaris is a Unix-derived Operating System produced by Sun Microsystems
    (\url{http://www.sun.com/software/solaris/}).
}}

\glossary{name={awk},description={AWK is a general purpose programming
    language that is designed for processing text-based data, and is
    available as a standard utility on all Unix systems.
}}
